### 2024-02-29 course/csic5011/2024/homework/homework03.tex
homework03

\begin{enumerate}
\item (James-Stein) Consider the following two-level Gaussian sampling process: let $\mu_i \sim \mathcal{N}(0,\sigma^2)$ ($i=1,\ldots, n$), and for each $i$, $x_i\sim \mathcal{N}(\mu_i,1)$. In other words, $\mu_i$ is sampled from a Gaussian prior $ p(\theta) = \mathcal{N}(0,\sigma^2)$ with unknown $\sigma^2$ and data $x_i$ is further sampled from Gaussian distribution $p(x|\theta)=\mathcal{N}(\mu_i,1)$. Show that 
\subitem[i.] the marginal $p(x) = \mathcal{N}(0, \sigma^2+1)$; 
\subitem[ii.] the posterior $p(\theta | x) = \mathcal{N}\left( \frac{\sigma^2 x}{\sigma^2+1}, \frac{\sigma^2}{\sigma^2+1} \right)$;
\subitem[iii.] the posterior mean gives
\[  \E[\theta | x] = \frac{\sigma^2 x}{\sigma^2+1} = \left( 1 - \frac{1}{\sigma^2+1} \right) x
\] 
which equals to by plugging in the unbiased estimate $\widehat{\sigma}^2 + 1 = \frac{1}{n-1} \sum_{i=1}^n x_i^2$,
\[  \widehat{\mu}^{JS} =  \left( 1 - \frac{n-1 }{\sum_{i=1}^n x_i^2} \right) x .
\] 
\item (Tweedie Formula) Consider a general prior $p(\theta)$ and the Gaussian likelihood $p(x|\theta) = \mathcal{N}(\theta,\sigma^2)$. Show that the posterior mean must be
\begin{equation} 
\E[\theta | x] = x + \sigma^2 \frac{d}{dx} \log p(x) 
\end{equation} 
which does NOT depends on the prior distribution $p(\theta)$, but just the gradient of score function of marginal ($\log p(x)$)!  This generalization of James-Stein estimator is called Tweedie Formula, which was recently found useful in denoising diffusion models and transformers.
\end{enumerate}

>>>>

\begin{enumerate}
\item (James-Stein) Consider the following two-level Gaussian sampling process: let $\mu_i \sim \mathcal{N}(M,A)$ ($i=1,\ldots, n$), and for each $i$, $x_i\sim \mathcal{N}(\mu_i,1)$. In other words, $\mu_i$ is sampled from a Gaussian prior $ p(\theta) = \mathcal{N}(M,A)$ with unknown $A,M$ and data $x_i$ is further sampled from Gaussian distribution $p(x|\theta)=\mathcal{N}(\mu_i,1)$. Show that 
\subitem[i.] the marginal $p(x) = \mathcal{N}(M, A+1)$; 
\subitem[ii.] the posterior $p(\theta | x) = \mathcal{N}\left( {\bf M}+B({\bf x}-{\bf M}), B\right)$ where $B=\frac{A}{A+1}$, ${\bf x}=(x_1,\ldots,x_n)^T$ and ${\bf M}=(M,\ldots, M)^T$ as vector;
\subitem[iii.] the posterior mean gives, for each $i=1,\ldots,n$
\[ \widehat{\mu}_i^{Bayes} = \E[\theta | x_i] = M + B(x_i-M) 
\] 
where the unbiased estimates of $M$ and $B$ give $\widehat{M}=\bar{x}=\frac{1}{n}\sum_{i=1}^n x_i$, $\widehat{B}=1-(n-3)/S$ with $S = \sum_{i=1}^n x_i^2$. This leads to the James-Stein estimator in the form 
\[  \widehat{\mu}^{JS}_i =  \bar{x} + \left(1- \frac{n-3 }{S} \right) (x_i - \bar{x}) .
\] 
\item (Tweedie Formula) Consider a general prior $p(\theta)$ and the Gaussian likelihood $p(x|\theta) = \mathcal{N}(\theta,\sigma^2)$. Show that the posterior mean must be
\begin{equation} 
\E[\theta | x] = x + \sigma^2 \frac{d}{dx} \log p(x) 
\end{equation} 
which does NOT depends on the prior distribution $p(\theta)$, but just the gradient of score function of marginal ($\log p(x)$)!  This generalization of James-Stein estimator is called Tweedie Formula, which was recently found useful in denoising diffusion models and transformers.
\end{enumerate}
